---
title: "Data Binning and Correlation"
output: 
    html_document:
        toc: TRUE
---

<style>
pre {
  overflow-x: auto;
}
pre code {
  word-wrap: normal;
  white-space: pre;
}
a {
  font-size: 16px;
}
</style>

```{r global options, include = FALSE}
knitr::opts_chunk$set(echo=TRUE, include = TRUE, warning=FALSE, message=FALSE)
```

---


* Aim: demonstrate how binning affects correlation 

---

### Simulation

* random data from normal distribution
* set.seed for reproducibility
* seed was purposely selected to have a weak association

```{r}
# set n
n <- 5000

# generate data
set.seed(285)

df <- data.frame(x = rnorm(n),
                 y = rnorm(n))

# order by x
df <- df[order(df$x),]
```


```{r fig.align="center", fig.height=6, fig.width=6}
# plot data
par(mfrow=c(1,1))

plot(y ~ x, data = df, main = "", pch=19, col = "#00000022", xlim = c(-4,4), ylim = c(-4,4))
legend("topleft", legend = paste("r =", round(cor(df$x,df$y),2)))
legend("topright", legend = paste("pval =", format(cor.test(df$x,df$y)$p.value, digits = 2, scientific = T)))
abline(lm(y ~x, data = df), col="red3", lwd=2)
```



* very weak assosication
* p-value is misleading (<0.05)

---

### Binning

* bin and average both x and y variables

```{r fig.align='center', fig.height=10, fig.width=10}

par(mfrow=c(2,2), oma=c(2,2,2,2))

# plot without binning
plot(y ~ x, data = df, main = "no binning", pch=19, col = "#00000022")
legend("topleft", legend = paste("r =", round(cor(df$x,df$y),2)))
legend("topright", legend = paste("pval =", format(cor.test(df$x,df$y)$p.value, digits = 2, scientific = T)))
abline(lm(y ~x, data = df), col="red3", lwd=2)


# iterate through different number of bins
for(n_bins in c(150,50,20)){
    x_binned <- sapply(1:n_bins, FUN = function(i){  mean(df$x[((i-1)*(n/n_bins)+1):((i)*(n/n_bins))])})
    y_binned <- sapply(1:n_bins, FUN = function(i){  mean(df$y[((i-1)*(n/n_bins)+1):((i)*(n/n_bins))])})
    
    plot(x_binned, y_binned, main = paste("n bins =", n_bins), pch=19, col = "#00000055")
    legend("topleft", legend = paste("r =", round(cor(x_binned,y_binned),2)))
    legend("topright", legend = paste("pval =", format(cor.test(x_binned,y_binned)$p.value, digits = 2, scientific = T)))
    abline(lm(y_binned ~ x_binned), col="red3", lwd=2)
}

```

* using fewer bins (more data points in one bin) the correlation becomes greater
* p-value changed slightly
* note that axis ranges are different (it makes slopes look also different)

---

```{r fig.align='center', fig.height=10, fig.width=10}

par(mfrow=c(2,2), oma=c(2,2,2,2))

# plot without binning
plot(y ~ x, data = df, main = "no binning", pch=19, col = "#00000022", xlim = c(-4,4), ylim = c(-4,4))
legend("topleft", legend = paste("r =", round(cor(df$x,df$y),2)))
legend("topright", legend = paste("pval =", format(cor.test(df$x,df$y)$p.value, digits = 2, scientific = T)))
abline(lm(y ~x, data = df), col="red3", lwd=2)


# iterate through different number of bins
for(n_bins in c(150,50,20)){
    x_binned <- sapply(1:n_bins, FUN = function(i){  mean(df$x[((i-1)*(n/n_bins)+1):((i)*(n/n_bins))])})
    y_binned <- sapply(1:n_bins, FUN = function(i){  mean(df$y[((i-1)*(n/n_bins)+1):((i)*(n/n_bins))])})
    
    plot(x_binned, y_binned, main = paste("n bins =", n_bins), pch=19, col = "#00000055", xlim = c(-4,4), ylim = c(-4,4))
    legend("topleft", legend = paste("r =", round(cor(x_binned,y_binned),2)))
    legend("topright", legend = paste("pval =", format(cor.test(x_binned,y_binned)$p.value, digits = 2, scientific = T)))
    abline(lm(y_binned ~ x_binned), col="red3", lwd=2)
}

```

* slopes are actually similar


---

### Conclusions

* binning falsifies weak association to strong!
* one has to be really careful with data binning


---

### Links

https://www.tandfonline.com/doi/abs/10.1080/09332480.2006.10722771

https://en.wikipedia.org/wiki/Ecological_fallacy

https://stats.stackexchange.com/questions/106016/is-using-deciles-to-find-correlation-a-statistically-valid-approach

https://statmodeling.stat.columbia.edu/2016/06/17/29400/


