---
title: "Determining NFR width by cross-correlation"
output: 
    html_document:
        toc: TRUE
---


<style>
a {
  font-size: 16px;
}
</style>


```{r global options, include = FALSE}
knitr::opts_chunk$set(echo=TRUE, include = TRUE, warning=FALSE, message=FALSE)
```

---

## Nucleosome Data by MNase-seq

* dyad coverages are centered at different kind of sites
    + transcription factor (TF) motif (PWM) positions
    + transcription start sites (TSS)
    + TSS +1 nucleosome
* goal is to estimate the distance between flanking nucleomes surrounding the center
* this can be called NFR (nucleosome "free" region) width
* (side note: one could subtract 147 bp (1 nucleosome) from this width)

---


## Method Description

* the method is inspired by Ocampo et al. 2016 and Chereji et al. 2018
* 2 Gaussian distributions mimic 2 flanking nucleosomes with known distance between them (theoretical pattern)
* for each site, the nucleosome signal is compared to the theoretical pattern (by cross-correlation)
* the actual distance for the given site is selected by maximizing the cross-correlation

---

```{r}
### helper functions ###

# generate Gaussian distribution
gaussmf <- function (x, sigma, mean) {
      height <- 1
      mfVals = (height * exp(-(((x - mean)^2)/(2 * (sigma^2)))))
      mfVals
}

# create NFR-like pattern
makeNFR <- function(d, beforeRef, afterRef){
      
      apply(sapply(c(-0.5,0.5), function(x) {gaussmf(seq(-beforeRef, afterRef), 40, x * d)}), 1, sum)
      
}

# scale between minimum and maximum
scaler <- function(x, probs = c(0.01,0.99)){
      (x - quantile(x, probs[1])) / (quantile(x, probs[2])-quantile(x, probs[1]))
}


### main function ###

# iterate through each row of the matrix
# use ccf function to find maximum correlation
# pick d which corresponds to the maximum correlation

calcNFRwidth <- function(my_mat,    
                          beforeRef = 200,
                          afterRef =  200,
                          mc.cores = 16){

    
    # here before and after is not necessarily symmetric (e.g. if matrix is not centered at NFR)
    my_mat_sub <- my_mat[, (ceiling(ncol(my_mat)/2)-beforeRef):(ceiling(ncol(my_mat)/2)+afterRef-1)]
    
    
    res <- parallel::mclapply(1:nrow(my_mat_sub), mc.cores = mc.cores, function(ridx){
        
        rx <- scaler(my_mat_sub[ridx, ])
        
        bestR <- 0
        spacingV <- NA
        shiftV <- NA
        
        total_width <- beforeRef+afterRef
        
        if(round(sd(rx), 3) != 0 & sum(is.na(rx)) == 0){
            
            for (d in 1:(total_width+100)) {
                
                # here before and after should be symmetric (left and right to center)
                y <-  makeNFR(d = d, beforeRef = total_width/2, afterRef = total_width/2)
                
                my_ccf <- ccf(rx,y, lag.max = 50, plot=FALSE)
                
                r <- max(my_ccf$acf)
                shift <- my_ccf$lag[my_ccf$acf == r]
                
                if (r > bestR) {
                    bestR <- r
                    shiftV <- shift
                    spacingV <- d
                }
            }
            
        }
        
        
        c(r = round(bestR, 2), width = spacingV, shift = shiftV)
        
    })                         
    
    df <- t(as.data.frame(res))
    rownames(df) <- rownames(my_mat_sub)
    return(df)

}

```



```{r echo=F}
### plotting functions
# not shown in doc
library(EBImage)

plotHeatmap <- function(mat,
                        df,
                        plot_title,
                        min_value = 0,
                        max_value = 0.6){
    
    mat_hm <- scaler(mat)
    
    mat_hm[mat_hm < min_value ] <- min_value
    mat_hm[mat_hm > max_value ] <- max_value
    
    my_order <- order(df[,2], decreasing = T)
    
    image(t(mat_hm[rev(my_order),]), 
          main = plot_title,
          col = colorRampPalette(rev(brewer.pal(n = 9, name ="RdYlBu")))(100),
          breaks = seq(min(mat_hm), max(mat_hm), length.out = 101),
          xaxt="n", yaxt="n", 
          useRaster = T)
    
    axis(side = 2, at = 0.5, labels = "NFR width", tick = F)
    
    axis(side = 1, at = seq(0,1, length.out = 5), 
         labels =  seq(-round(ncol(mat_hm)/2000), round(ncol(mat_hm)/2000), length.out = 5 )*1000)
    axis(side = 4, at =  seq(0,1,0.1), labels = round(quantile(df[,2], na.rm = T, probs = seq(0,1,0.1))), las=2, cex.axis=0.8)
}


plotExamples <- function(examples){
    
    total_width <- beforeRef+afterRef

    for(i in seq_along(examples)){
    
    sidx <- examples[i]
    
    if(is.na(sidx)){next()}
    
    my_pattern <- makeNFR(d = df[sidx,2], beforeRef = total_width/2, afterRef = total_width/2)
    
    plot(df[sidx,3]:(length(my_pattern)+df[sidx,3]-1), my_pattern, type="l", xlab = "", xaxt ="n", ylab="Density", main = "")
    axis(side = 1, at = seq(df[sidx,3],(length(my_pattern)+df[sidx,3]-1), 100), labels = seq(-(total_width/2), (total_width/2), 100))
    
    mtext(paste("corr =", df[sidx,1],"width =", df[sidx,2], "shift =", df[sidx,3]), side = 3, line = 1, cex = 0.8)
    
    lines(scaler(my_mat[sidx,(ceiling(ncol(my_mat)/2)-beforeRef):(ceiling(ncol(my_mat)/2)+afterRef-1)]), col="red3")
    
    if(i %% 8 == 0){
        mtext(text = "Examples", side = 3, outer = T, font = 2)
        
        par(fig = c(0,1,0,1), mar=c(0,0,0,0), new = TRUE)
        plot(0,0, xlab="", ylab="", xaxt="n", yaxt="n", bty="n", type="n" )
        legend("center", legend = c("Actual", "Theoretical"), 
               horiz = TRUE, pch=19, col= c("red3", "black"), cex=1.25, bty="n")
    }    
    
}
}

```


---

## Budding Yeast Data #1

* in vivo MNase-seq from yeast (S.cerevisae)
* centered at putative TF sites (based on PWM)
* single end data: all fragments are considered


```{r}
# load site centered matrix
load("data_NFRwidth/smooth.mat.TF.yeast.invivo.MNase.rda")
```



```{r}
# use +/- 200 bp from the center
beforeRef <- 200 
afterRef  <- 200
```


```{r}
file_path <- paste0("data_NFRwidth/NFRwidth.TF.yeast.invivo.MNase.set",beforeRef,"_",afterRef,".txt")

if(file.exists(file_path)){
    
    #read NFR data (pre-calculated)
    df <- read.table(file_path, row.names = 1, header = T)
    
} else {
    
    # run analysis
    df <- calcNFRwidth(my_mat = my_mat,
                    beforeRef = beforeRef,
                    afterRef = afterRef)
    
    write.table(df, file = file_path, sep = "\t", row.names = T, col.names = NA)
}

head(df)
```

* the greater the correlation the more similar is the actual pattern to the theoretical
* width indicates the NFR width
* shift indicates how far the signal had to be shifted from the center to get the max correlation

---

#### Heatmap Visualization

```{r fig.align="center", fig.height=6, fig.width=4}
# heatmap  ordered by NFR width
library(RColorBrewer)

par(mfrow=c(1,1), oma=c(2,0,0,1))

plotHeatmap(mat = my_mat, df = df, plot_title = "Nucleosomes centered at TF")
```

* The order on the heatmap seems to be ok

---

#### Exploring the Values

```{r fig.align="center", fig.height=4, fig.width=10}
# check distributions
par(mfrow=c(1,3), oma=c(2,0,0,0))

hist(df[,1], breaks = 500, main = "NFR correlation", xlab="")
hist(df[,2], breaks = 250, main = "NFR width", xlab="")
hist(df[,3], breaks = 500, main = "NFR shift", xlab="")
```

* the majority of the width values are between 200-400
* there are some values at the extremes (1, 500)

---

#### Example Sites

```{r fig.align="center", fig.height=6, fig.width=10}
# choose examples by interesting values

examples <- c(which(df[,2] > 0   & df[,2] <= 1 & df[,1] > 0.9)[1],
              which(df[,2] > 165 & df[,2] <= 170 & df[,1] > 0.9)[1],
              which(df[,2] > 295 & df[,2] <= 305 & df[,1] > 0.9)[1],
              which(df[,2] > 499 & df[,2] <= 500 & df[,1] > 0.6)[1],
              which(df[,2] > 0   & df[,2] <= 1 & df[,1] < 0.7)[1],
              which(df[,2] > 165 & df[,2] <= 170 & df[,1] < 0.7)[1],
              which(df[,2] > 295 & df[,2] <= 305 & df[,1] < 0.7)[1],
              which(df[,2] > 499 & df[,2] <= 500 & df[,1] < 0.6)[1])

par(mfrow=c(2,4), oma=c(0,0,2,0))

plotExamples(examples = examples)
```

* width = 1 (no distance) means that a single nucleosome is sitting at the center
* width = 500 suggests a nucleosome is at one side only (not a real NFR?)
* lower correlation values might give a reasonable width although the pattern is more noisy




---

## Budding Yeast Data #2

* in vivo MNase-seq from yeast (S.cerevisae)
* paired end data: mono-nuclesomal fragments are selected
* centered at TSS +1 nucleosome (which was determined previously by Chereji et al. 2018)
* this creates a problem that the matrix is not centered at the NFR
* (Side note: I could have used the NFR center positions from Chereji et al.)
* (Side note2: Chereji et al. also use a similar method where they first map the +1 and -1 nucleosomes)
* one goal is here to test different window sizes left and right


```{r}
# load site centered matrix
load("data_NFRwidth/smooth.mat.plusOne.yeast.invivo.MNase.rda")
```


```{r}
# settings are different window sizes
# windows are asymmetric as the signal is centered at the +1 nucleosome
settings <- list(c(450,50),
                 c(400,50),
                 c(350,50),
                 c(300,50))

for(i in seq_along(settings)){
    
    beforeRef <- settings[[i]][1] 
    afterRef  <- settings[[i]][2] 
        
    file_path <- paste0("data_NFRwidth/NFRwidth.plusOne.yeast.invivo.PE.MNase.set",beforeRef,"_",afterRef,".txt")

    if(file.exists(file_path)){
        
        #read NFR data (pre-calculated)
        df <- read.table(file_path, row.names = 1, header = T)
        
        assign(paste0("df",i), df)
        
    } else {
        # run analysis
        df <- calcNFRwidth(my_mat = my_mat,
                            beforeRef = beforeRef,
                            afterRef = afterRef)
        write.table(df, file = file_path, sep = "\t", row.names = T, col.names = NA)

        assign(paste0("df",i), df)
    }
}
```

---

#### Heatmap Visualization

```{r fig.align="center", fig.height=6, fig.width=12}
# heatmap  ordered by NFR width
library(RColorBrewer)

par(mfrow=c(1,4), oma=c(1,0,1,1))

# iterate through settings
for(i in seq_along(settings)){
    
    plotHeatmap(mat = my_mat, df = get(paste0("df",i)), 
            plot_title = paste0("beforeRef = ",settings[[i]][1],"; afterRef =",settings[[i]][2]))
}

title(main = "Nucleosomes centered at TSS+1 nucleosome", outer = TRUE)
```

* 1st window does not capture -1 properly (there is a small nuclesome in between)
* 3rd window looks the most reasonable

---

#### Exploring the Values

```{r fig.align="center", fig.height=4, fig.width=10}
# check distributions
par(mfrow=c(1,3), oma=c(2,0,0,0))

df <- df3
beforeRef = 350; afterRef = 50

hist(df[,1], breaks = 500, main = "NFR correlation", xlab="")
hist(df[,2], breaks = 250, main = "NFR width", xlab="")
hist(df[,3], breaks = 500, main = "NFR shift", xlab="")
```

* there is a shoulder at 400 (probably because window size is limited?)

---

#### Example Sites

```{r fig.align="center", fig.height=6, fig.width=10}
# choose examples by interesting values

examples <- c(which(df[,2] > 0   & df[,2] <= 1 & df[,1] > 0.8)[1],
              which(df[,2] > 150 & df[,2] <= 170 & df[,1] > 0.8)[1],
              which(df[,2] > 295 & df[,2] <= 305 & df[,1] > 0.9)[1],
              which(df[,2] > 499 & df[,2] <= 500 & df[,1] > 0.6)[2],
              which(df[,2] > 0   & df[,2] <= 1 & df[,1] < 0.7)[1],
              which(df[,2] > 165 & df[,2] <= 170 & df[,1] < 0.7)[1],
              which(df[,2] > 295 & df[,2] <= 305 & df[,1] < 0.7)[1],
              which(df[,2] > 499 & df[,2] <= 500 & df[,1] < 0.6)[1])

par(mfrow=c(2,4), oma=c(0,0,2,0))

plotExamples(examples = examples)
```

* the profiles with high correlation look quite ok

---

## Compare to Chereji etal. data

* Chereji et al. used a different method
* they first determine -1 and +1 nucleomes positions
* they use cross-correlation to refine those positions
* they detemine NFR width (NDR width) as the distance between -1 and +1 
* they substract 147 bp (1 nucleosome)

```{r fig.align="center", fig.height=6, fig.width=6}
# read published data
NFR <- read.table("data_NFRwidth/13059_2018_1398_MOESM2_ESM.txt", sep = "\t", header = T, stringsAsFactors = F)

library(LSD)
par(mfrow=c(1,1), oma=c(2,2,0,2))


# merge datasets
df_NFR <- merge(df, NFR, by.x = "row.names", by.y = "ORF")

# add 147 for comparison
heatscatter(df_NFR$width, df_NFR$NDR.Width+147,
            xlab = "cross-correlation method", ylab = "Chereji method",
            main = "Comparison", pch=19, cex=0.6,
            xlim = c(100,600), ylim = c(100,600))

```

* there is a good match for NFRs between 200-400
* the limit at 400 is visible here as well

---

#### Heatmap Visualization

```{r fig.align="center", fig.height=6, fig.width=6}
# heatmap  ordered by NFR width
par(mfrow=c(1,2), oma=c(1,0,1,1))

my_mat <- my_mat[rownames(my_mat) %in% df_NFR$Row.names,]
my_mat <- my_mat[order(rownames(my_mat)),]

identical(as.character(df_NFR$Row.names), rownames(my_mat))

df <- df_NFR[,c(2,3,4)]
rownames(df) <- df_NFR$Row.names

plotHeatmap(mat = my_mat, df = df, plot_title = "cross-correlation method")


df <- df_NFR[,c(2,14,4)]
rownames(df) <- df_NFR$Row.names

plotHeatmap(mat = my_mat, df = df, plot_title = "Henikoff method")

title(main = "Nucleosomes centered at TSS+1 nucleosome", line = -0.5, outer = TRUE)
```

Conclusions:

* NFR width up to the window limit matches published results
* longer NFRs are not captured well with this method
* using larger window sizes is not an option (small nucleosome appears in the middle)

* the method can be still useful in the first example when centered at TF motif
* in this case the nucleosome might be exactly at the TF center (which results in a width=1)


---

## Data Source

The first dataset was generated by Elisa Oberbeckmann (LMU, BMC, Korber group)

The second dataset was generated by Ashish Singh (LMU, BMC, Müller-Planitz group)


---


## References


The ISW1 and CHD1 ATP-dependent chromatin remodelers compete to set nucleosome spacing in vivo.
Ocampo J, Chereji RV, Eriksson PR, Clark DJ.
Nucleic Acids Res. 2016 Jun 2;44(10):4625-35. doi: 10.1093/nar/gkw068. Epub 2016 Feb 9.
PMID: 26861626 

Precise genome-wide mapping of single nucleosomes and linkers in vivo.
Chereji RV, Ramachandran S, Bryson TD, Henikoff S.
Genome Biol. 2018 Feb 9;19(1):19. doi: 10.1186/s13059-018-1398-0.
PMID: 29426353 
