---
title: "K-Means Clustering - Female/Male Embryos"
output: 
    html_document:
        toc: TRUE
---


<style>
a {
  font-size: 16px;
}
</style>


```{r global options, include = FALSE}
knitr::opts_chunk$set(echo=TRUE, include = TRUE, warning=FALSE, message=FALSE)
```

---

## Embryo RNA-seq Data

* time course experiment of individual Drosophila embryos
* the time spans up to 12 hours in development
* during this time key sex determination genes get activated
* goal: distinguish which embryo is female or male

---

### Pre-processing

* alignment and count table was generated using STAR

```{bash eval=F}
# do not run here
module load ngs/STAR/2.5.3a
module load ngs/samtools

### STAR index directory ###
STAR_INDEX="../genome/STAR_dmel-all-chromosome-r6.17"

### GTF file ###
GTF_FILE="../genome/dmel-all-r6.17.gtf"

# STAR run
STAR \
--runThreadN 8 \
--readFilesCommand gunzip -c \
--quantMode GeneCounts \
--genomeDir ${STAR_INDEX} \
--sjdbGTFfile ${GTF_FILE} \
--readFilesIn ${FILEBASE}_1.txt.gz ${FILEBASE}_2.txt.gz \
--outFileNamePrefix ${FILEBASE}. \
--outSAMtype BAM SortedByCoordinate \
--limitBAMsortRAM 5000000000 \
--outFilterMultimapNmax 1
```

### Annotation

* annotation is derived from the same GTF (loaded as txdb)

```{r}
library(AnnotationDbi)
library(GenomicFeatures)

txdb = loadDb("data_Kmeans/txdb")

my_exons <- exons(txdb)

exons.list.per.gene <- exonsBy(txdb, by="gene")
exonic.gene.sizes <- sum(width(reduce(exons.list.per.gene))) / 1000

head(exonic.gene.sizes)
```


### Count Table

```{r}
# read data from text file
my_counts <- read.table("data_Kmeans/count_table.txt", header = T, row.names = 1)

my_counts[1:6,1:8]

dim(my_counts)
```

---

### Filtering and Normalization

* TPM normalization makes sure that samples and genes are comparable
    + first normalize by exonic gene length
    + second normalize by the sum of length corrected counts 

```{r}
# function to calculate TPM (see reference below)
countToTpm <- function(counts, effLen, scaler=1e6)
{
      rate <- log(counts) - log(effLen)
      denom <- log(sum(exp(rate)))
      exp(rate - denom + log(scaler))
}

```


```{r}
# select genes
my_counts_Genes <- my_counts[grep("FBgn", rownames(my_counts)),]

# filter low counts
my_filter <- apply(my_counts_Genes, 1, function(x) length(x[x>20]) >= ncol(my_counts_Genes)/12)

my_counts_Filtered <- my_counts_Genes[my_filter,]

# match filtered genes and exonic gene sizes
my_genes_Filtered <- intersect(rownames(my_counts_Filtered), names(exonic.gene.sizes))

exonic.gene.sizes <- exonic.gene.sizes[names(exonic.gene.sizes) %in% my_genes_Filtered]
exonic.gene.sizes <- exonic.gene.sizes[order(names(exonic.gene.sizes))]

my_counts_Filtered <- my_counts_Filtered[rownames(my_counts_Filtered) %in% my_genes_Filtered,]
my_counts_Filtered <- my_counts_Filtered[order(rownames(my_counts_Filtered)),]

# check whether names are identical
stopifnot(identical(rownames(my_counts_Filtered), names(exonic.gene.sizes)))

# calculate TPM
my_TPM <- apply(my_counts_Filtered, 2, FUN = function(x){countToTpm(x, exonic.gene.sizes)})

# take log
log2_TPM <- log2(my_TPM+0.5)

log2_TPM[1:6,1:6]

dim(log2_TPM)
```

---

# K-means Clustering

### Clustering with k=3

* first embryo is before sex determination is initiated
* that´s why we set k = 3 (early, female and male)

```{r}
library(org.Dm.eg.db)

# select well known sex-specific marker genes
my_favorite_genes <- c("Sxl", "msl-2" ,"lncRNA:roX1", "lncRNA:roX2")
names(my_favorite_genes) <- mapIds(org.Dm.eg.db, my_favorite_genes, "FLYBASE", keytype="SYMBOL", multiVals="first")

set.seed(99)

# k-means with k=3
kmeans_clusters <- kmeans(apply(log2_TPM[ rownames(log2_TPM) %in% names(my_favorite_genes), ], 1, 
                         function(x){ (x-mean(x))/sd(x) }),3, nstart=25, iter.max=1000)

my_sexes <- factor(as.integer(kmeans_clusters$cluster))

head(my_sexes)
```

### Plot Marker Genes

* plot TPM values to see sex-specific pattern

```{r echo=F}
# helper function for plotting

plotDots <- function(my_data, 
                     my_title,
                     color_palette,
                     color_groups,
                     conditions,
                     point_size = 1.5,
                     ylims = c(0, 15),
                     x_label = NULL){
      
    par(cex.main = 1.5)
      
      grouped_data = data.frame(exp = as.numeric(my_data),
                                group =  as.numeric(conditions))
      
      plot(grouped_data$group + runif(length(grouped_data$group), -0.05, 0.05),
           grouped_data$exp,
           pch=19, cex = point_size, ylim =  ylims,
           xaxt = "n", xlab = "", ylab = "",
           main= my_title, col =  color_palette[color_groups])
      
      if(!(is.null(x_label))){
            axis(side = 1, at = seq_along(unique(grouped_data$group)), labels = x_label)
            
      } else {
            axis(side = 1, labels = FALSE)
            
      }
     
}
```


```{r fig.align="center", fig.height=4, fig.width=10}
# conditions
my_conditions <- factor(substr(colnames(log2_TPM),1,2))

# setup color
my_color_palette <- c("#999999", "#d6604d", "#4393c3")

par(mfrow=c(1,4), oma=c(4,2,0,2))

# iterate through marker genes
for(i in seq_along(my_favorite_genes)){
    
    plotDots(log2_TPM[rownames(log2_TPM) == names(my_favorite_genes)[i],], 
                     my_title = my_favorite_genes[i], 
                     color_palette = my_color_palette, 
                     color_groups = my_sexes, 
                     conditions = my_conditions, 
                     point_size = 0.8)
}

mtext(text = "Stages", side = 1, line = -2, outer = TRUE)
mtext(text = "log2 TPM", side = 2, outer = TRUE)

par(fig = c(0,1,0,1), mar=c(0,0,0,0), new = TRUE)
plot(0,0, xlab="", ylab="", xaxt="n", yaxt="n", bty="n", type="n" )
legend("bottomright", legend = c("Early", "Female", "Male"), pch = 19, col = my_color_palette, horiz = T)

```






Conclusions:

* Expected sex-specific pattern:
    + Sxl is female, whereas msl-2 male specifc
    + roX1 is unspecfic until it becomes male specific
    + roX2 expression is activated later only in males
*  Using RNA-seq expression of known sex-specific genes one can easily distinguish between female and male embryos

---

# Data Source

The  dataset was generated by Tamas Schauer (LMU, BMC, Becker group)


---


# References

Progressive dosage compensation during Drosophila embryogenesis is reflected by gene arrangement.
Prayitno K, Schauer T, Regnard C, Becker PB.
EMBO Rep. 2019 Aug;20(8):e48138. doi: 10.15252/embr.201948138. Epub 2019 Jul 9.
PMID: 31286660

TPM calculation:
https://haroldpimentel.wordpress.com/2014/05/08/what-the-fpkm-a-review-rna-seq-expression-units/

